---
title: "Package architecture"
author: "Pablo Fuenzalida"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```


## Start here if you are new

**Sharkipedia** ([sharkipedia.org](https://www.sharkipedia.org)) is a website/database of shark and ray **life-history traits** and **population trends**, each tied to a literature reference.

**sharkipediaR** downloads public species pages, reads the HTML tables, and gives you clean R data frames. You do not need to know how the website is built.


## How the package is organised (one sentence)

Each `sp_*()` function **fetches a page → parses tables → cleans names → checks columns → returns a tibble**. That split keeps the code maintainable when the website changes.


---

## `sp_species_urls()` — find species web addresses

**What it does:** Reads the [species index](https://www.sharkipedia.org/species) and returns names + URLs.


**How to use:** Start here when you need a species list. Default = first page only (polite). Use `all_pages = TRUE` for the full catalogue (slow).


**Example:**

```{r sp-urls, eval = FALSE}
suppressPackageStartupMessages(library(tidyverse))
library(sharkipediaR)

idx <- sp_species_urls()
idx %>% slice_head(n = 5)
```


---

## `sp_search()` — search by name

**What it does:** Filters the index for partial scientific names (e.g. `"Carcharh"`).


**How to use:** Pass your own index if you already called `sp_species_urls(all_pages = TRUE)` — avoids re-downloading.


**Example:**

```{r sp-search, eval = FALSE}
sp_search("Galeocerdo", index = idx)
```


---

## `sp_species()` — taxonomy for one species

**What it does:** Returns one row: species name, order, family, etc., plus when/where it was downloaded.


**Example:**

```{r sp-species, eval = FALSE}
sp_species("Carcharodon carcharias")
```


---

## `sp_traits()` — life-history measurements

**What it does:** All trait rows from a species page (age at maturity, length, reproduction, …) in **long format** — one row per measurement.


**How to use:** Pass one name or a **vector** of names for batch downloads.


**Example:**

```{r sp-traits, eval = FALSE}
traits <- sp_traits("Carcharodon carcharias")
traits %>%
  filter(trait_name == "Linf") %>%
  mutate(value = as.numeric(value))
```


---

## `sp_trends()` — population trends through time

**What it does:** Extracts year–value series from trend charts on the species page (the same data behind Sharkipedia’s plots).


**How to use:** Filter by `reference` to match a paper (see the **Ecological workflows** tutorial for `reid2011`).


**Example:**

```{r sp-trends, eval = FALSE}
trends <- sp_trends("Carcharodon carcharias")
trends %>%
  filter(reference == "reid2011") %>%
  ggplot(aes(year, value)) +
  geom_line()
```


---

## `sp_references()` — literature links on a species page

**What it does:** Lists reference IDs (e.g. `reid2011`) and URLs cited on that species page.


**Example:**

```{r sp-refs, eval = FALSE}
sp_references("Carcharodon carcharias")
```


---

## `example_carcharhinus()`, `example_reid2011_trends()`, `example_white_shark_trends()`

**What they do:** Frozen Sharkipedia data for tutorials and testing — **no internet** required.


**Example:**

```{r examples}
suppressPackageStartupMessages(library(tidyverse))
library(sharkipediaR)

example_reid2011_trends() %>% count(species)
```


---

## `fetch_page()` — advanced

**What it does:** Downloads raw HTML only. Use this if you extend the package; most people should use `sp_*()` instead.


**Example:**

```{r fetch, eval = FALSE}
# fetch_page("https://www.sharkipedia.org/species/carcharodon-carcharias")
```


---

## Behind the scenes (optional reading)

| Step | File | Role |
|------|------|------|
| Download | `fetch.R` | Polite HTTP, rate limit, cache |
| Read HTML | `parse.R` | Find tables and trend JSON |
| Tidy | `clean.R` | snake_case columns, provenance |
| Check | `validate.R` | Required columns present |

You never need to call these directly unless you are contributing to **sharkipediaR**.


---

## Be a good citizen

- Do not hammer the site with parallel requests.
- Keep `source_url` and `retrieved_at` in your outputs.
- Cite Sharkipedia and the original papers (e.g. Reid et al. 2011).
