---
title: "Package architecture and function reference"
author: "Pablo Fuenzalida"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package architecture and function reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

```{r load}
library(sharkipediaR)
library(dplyr)

vignette_fixture <- function(name) {
  path <- system.file("fixtures", name, package = "sharkipediaR")
  if (!nzchar(path)) {
    cli::cli_abort("Fixture not found: {name}")
  }
  path
}

doc <- xml2::read_html(vignette_fixture("carcharhinus_acronotus.html"))
index_doc <- xml2::read_html(vignette_fixture("species_index_page1.html"))
source_url <- "https://www.sharkipedia.org/species/carcharhinus-acronotus"
retrieved_at <- as.POSIXct("2026-05-25 12:00:00", tz = "UTC")
```

## Design philosophy

**sharkipediaR** follows a strict pipeline:

```
HTTP retrieval  →  HTML parsing  →  cleaning  →  validation  →  user-facing tibble
(fetch.R)         (parse.R)        (clean.R)     (validate.R)   (sp_*.R)
```

No step is allowed to “do everything.” That separation keeps the package maintainable when Sharkipedia’s HTML evolves.

```{r diagram, echo = FALSE, fig.cap = "Data flow for sp_traits()"}
par(mar = c(0, 0, 0, 0))
plot.new()
text(0.5, 0.85, "sp_traits(species)", cex = 1.2, font = 2)
arrows(0.5, 0.78, 0.5, 0.72, length = 0.1)
text(0.5, 0.68, "resolve_species_url()  →  fetch_page()", cex = 0.95)
arrows(0.5, 0.62, 0.5, 0.56, length = 0.1)
text(0.5, 0.52, "parse_traits_tables()", cex = 0.95)
arrows(0.5, 0.46, 0.5, 0.40, length = 0.1)
text(0.5, 0.36, "clean_traits()  →  validate_traits()", cex = 0.95)
```

---

# `R/constants.R`

| Object / function | Role |
|-----------------|------|
| `.sharkipedia_env` | Package-private environment storing last request time for rate limiting |
| `sharkipedia_base_url()` | Returns `https://www.sharkipedia.org` |
| `sharkipedia_user_agent()` | Identifies the client in HTTP headers (version + project URL) |

These are internal but underpin polite scraping.

---

# `R/utils.R`

## `rate_limit_pause(min_gap = 0.5)`

Waits until at least `min_gap` seconds have elapsed since the last request, plus a small random jitter (`runif(0, 0.3)`). Called automatically by `fetch_page()`.

## `species_name_to_slug(name)`

Converts `"Carcharhinus acronotus"` → `"carcharhinus-acronotus"` (lowercase, spaces to hyphens).

## `species_name_to_url(name)`

Builds the full species page URL from a scientific name.

```{r utils-url}
sharkipediaR:::species_name_to_url("Carcharhinus acronotus")
```

## `normalize_sharkipedia_url(url)`

Accepts:

- full URLs,
- relative paths (`/species/...`),
- bare scientific names.

## `resolve_species_url(x)`

Length-1 resolver used by all `sp_*()` functions. Errors if `length(x) != 1`.

## `ensure_species_vector(x)`

Trims, de-duplicates, and validates a character vector for batch functions.

## `read_html_fixture(path)`

Reads local HTML for tests (internal).

## `get_fetch_page(cache = TRUE)`

Returns either `fetch_page` or memoised `fetch_page_memoised` depending on the `cache` argument passed to user functions.

---

# `R/fetch.R`

## `fetch_page(url, quiet = TRUE)` — **exported**

**Responsibilities only:** HTTP, rate limit, retries, return `xml_document`.

| Step | Implementation |
|------|----------------|
| URL normalisation | `normalize_sharkipedia_url()` |
| Politeness | `rate_limit_pause()` |
| Request | `httr2::request()` + user agent + `req_retry(max_tries = 3)` |
| Error handling | Abort on status ≥ 400 with `cli` message |
| Body | `httr2::resp_body_html()` → **xml2** document |

```{r fetch-demo, eval = FALSE}
doc_live <- fetch_page("Carcharhinus acronotus", quiet = FALSE)
class(doc_live)
```

## `fetch_page_memoised`

`memoise::memoise(fetch_page)` — identical URLs in one R session return the cached document without a new HTTP call.

---

# `R/parse.R`

Parsing functions **never** perform HTTP. They only read `xml_document` objects.

## `parse_species_index(doc)`

**Used by:** `sp_species_urls()`

- Selector: `a[href^='/species/']`
- Excludes CSV export links (`*.csv` slugs)
- Returns `species`, `slug`, `url`

```{r parse-index}
idx <- sharkipediaR:::parse_species_index(index_doc)
head(idx, 4)
```

## `parse_index_last_page(doc)`

Reads Bulma pagination links (`.pagination-link`) and returns the maximum page number — used when `sp_species_urls(all_pages = TRUE)`.

## `parse_taxonomy(doc)`

**Used by:** `sp_species()`

- Species name: `h1.title`
- Ranks: `div.columns > div.column:first-child > p` with labels `Superorder:`, `Subclass:`, `Order:`, `Family:`

```{r parse-tax}
sharkipediaR:::parse_taxonomy(doc)
```

## Trait table helpers

| Function | Role |
|----------|------|
| `trait_table_headers()` | Reads `<thead><th>` text |
| `is_trait_table()` | `TRUE` if headers include `"Name"` |
| `preceding_trait_group()` | XPath `preceding::h4[contains(@class,'subtitle')][1]` for trait class |

## `parse_traits_tables(doc)`

**Used by:** `sp_traits()` via `fetch_species_traits()`

1. Find all `table.table` elements.
2. Keep tables whose headers match Sharkipedia’s trait schema.
3. `rvest::html_table()` for body rows.
4. Attach `trait_group` from the nearest preceding `h4`.

```{r parse-traits-raw}
raw_traits <- sharkipediaR:::parse_traits_tables(doc)
names(raw_traits)
head(raw_traits, 3)
```

Note: column names are still **PascalCase** (`Name`, `Value`, …) at this stage.

## `decode_react_props(props)`

Decodes HTML entities in `data-react-props` JSON and parses with **jsonlite**. Internal helper for trends.

## `parse_trends_tables(doc)`

**Used by:** `sp_trends()`

1. Locate the first table after the `h3` heading containing `"Trends"`.
2. For each `<tr>`:
   - `location`, `unit`, `reference` from table cells
   - `trend_id` / `trend_url` from the “Details” link
   - `observations` matrix from `div[data-react-props]`
3. Expand to one row per **year × value**.

```{r parse-trends-raw}
raw_trends <- sharkipediaR:::parse_trends_tables(doc)
raw_trends %>%
  count(location, trend_id) %>%
  head(5)
```

## `parse_references(doc)`

Collects unique `a[href^='/references/']` from trait and trend tables.

```{r parse-refs}
sharkipediaR:::parse_references(doc) %>% head(5)
```

## `link_text_or_href(node)`

Utility for anchor text vs. `href` basename (internal).

---

# `R/clean.R`

Cleaning standardises names, types, and provenance.

## `standardize_trait_columns(df)`

Maps Sharkipedia headers to snake_case:

| HTML | R column |
|------|----------|
| Name | trait_name |
| Value | value |
| Standard | standard |
| ValueType | value_type |
| Sex | sex |
| Location | location |
| Reference | reference |

## `clean_traits(df, species, source_url, retrieved_at)`

- Renames columns via `standardize_trait_columns()`
- Adds `species`, `source_url`, `retrieved_at` (UTC POSIXct)
- `str_squish()` + `na_if("", .)` on character fields
- Strips `/references/` prefixes from `reference`

```{r clean-traits}
traits_clean <- sharkipediaR:::clean_traits(
  raw_traits,
  species = "Carcharhinus acronotus",
  source_url = source_url,
  retrieved_at = retrieved_at
)
names(traits_clean)
```

## `clean_trends(df, species, source_url, retrieved_at)`

- Coerces `year` to integer, keeps `value` as double
- Adds species and provenance columns
- Squashes whitespace on metadata fields

```{r clean-trends}
trends_clean <- sharkipediaR:::clean_trends(
  raw_trends,
  species = "Carcharhinus acronotus",
  source_url = source_url,
  retrieved_at = retrieved_at
)
summary(trends_clean$year)
```

---

# `R/validate.R`

Validation is the last gate before data reach the user.

## `validate_traits(df)`

- **Required columns:** `species`, `trait_name`, `value`, `source_url`, `retrieved_at`
- **Warns** (does not error) on zero rows — page structure may have changed

## `validate_trends(df)`

- **Required columns:** `species`, `year`, `value`, `source_url`, `retrieved_at`
- Same empty-data warning pattern

```{r validate}
sharkipediaR:::validate_traits(traits_clean) %>% nrow()
sharkipediaR:::validate_trends(trends_clean) %>% nrow()
```

---

# `R/sp_species_urls.R`

## `sp_species_urls(all_pages = FALSE, max_pages = NULL, cache = TRUE)`

| Argument | Behaviour |
|----------|-----------|
| `all_pages = FALSE` | Only page 1 of `?all=true` index (~20 species) |
| `all_pages = TRUE` | Reads pagination from page 1, loops all pages |
| `max_pages` | Caps pagination when exploring |
| `cache` | Uses memoised `fetch_page` |

Returns deduplicated tibble: `species`, `slug`, `url`.

```{r sp-urls, eval = FALSE}
sp_species_urls()
sp_species_urls(all_pages = TRUE, max_pages = 3)
```

---

# `R/sp_search.R`

## `sp_search(query, index = NULL, all_pages = FALSE, cache = TRUE)`

1. Normalises `query` via `ensure_species_vector()`.
2. Builds or accepts `index` tibble.
3. `grepl(..., fixed = TRUE)` on lowercased `species` and `slug`.
4. `distinct(url)`.

Best practice: `idx <- sp_species_urls(all_pages = TRUE)` once, then `sp_search("rhincodon", index = idx)`.

---

# `R/sp_species.R`

## `sp_species(species, cache = TRUE)`

End-user taxonomy extractor. Returns **one row**:

`species`, `superorder`, `subclass`, `order`, `family`, `source_url`, `retrieved_at`

```{r sp-species-demo}
ex <- example_carcharhinus()
ex$species_meta
```

---

# `R/sp_traits.R`

## `sp_traits(species, cache = TRUE)`

Public API for traits.

- **One species:** calls `fetch_species_traits()`.
- **Many species:** `purrr::map_dfr()` with column `species_input`.

## `fetch_species_traits(species, cache = TRUE)` — internal

Orchestrates the full trait pipeline documented above.

```{r full-pipeline-traits}
# Equivalent to fetch_species_traits() without HTTP:
traits_clean <- sharkipediaR:::validate_traits(traits_clean)
nrow(traits_clean)
```

---

# `R/sp_trends.R`

## `sp_trends(species, cache = TRUE)`

Public API for population trends (long format).

## `fetch_species_trends(species, cache = TRUE)` — internal

Same pattern as traits: fetch → parse → clean → validate.

```{r full-pipeline-trends}
nrow(trends_clean)
```

---

# `R/sp_references.R`

## `sp_references(species, cache = TRUE)`

Fetches species page, runs `parse_references()`, adds `species`, `source_url`, `retrieved_at`.

Use to build a reference lookup table before joining to traits/trends.

---

# `R/example_data.R`

## `example_carcharhinus()` — **exported**

Loads `inst/extdata/carcharhinus_acronotus.rds` — a list with pre-parsed components used in all vignettes.

Rebuild after fixture updates:

```{r rebuild, eval = FALSE}
source("data-raw/build-vignette-data.R")
```

---

# End-to-end manual pipeline (advanced users)

Reproduce `sp_traits()` without calling it:

```{r manual-pipeline}
url <- sharkipediaR:::species_name_to_url("Carcharhinus acronotus")
# doc <- fetch_page(url)   # live
# raw <- parse_traits_tables(doc)
# out <- validate_traits(clean_traits(raw, "Carcharhinus acronotus", url))
# Identical structure to example_carcharhinus()$traits
```

This is the extension point if Sharkipedia adds a JSON API later: swap `fetch_page()` + `parse_*()` implementations while keeping `sp_traits()` stable.

---

# Testing and pkgdown

- **Unit tests** use HTML fixtures in `tests/testthat/fixtures/` (no network).
- **Vignettes** use `example_carcharhinus()` for plots and tables.
- **pkgdown** picks up vignettes automatically after `pkgdown::build_site()`.

See **Getting started** and **Ecological workflows** for ggplot demonstrations and scientific motivation.
