---
title: "Ecological workflows with Sharkipedia data"
author: "Pablo Fuenzalida"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ecological workflows with Sharkipedia data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

```{r load}
library(sharkipediaR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(plotly)

ex <- example_carcharhinus()
traits <- ex$traits
trends <- ex$trends
meta <- ex$species_meta

theme_shark <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey30", size = 10),
      strip.text = element_text(face = "bold")
    )
}

# Wrap ggplot objects for interactive tooltips in HTML vignettes / pkgdown
as_interactive <- function(p) {
  plotly::ggplotly(p, tooltip = "all") %>%
    plotly::layout(hovermode = "closest")
}
```

## The scientific case

Chondrichthyans are among the most threatened vertebrate lineages. Comparative traits (maturity, growth, fecundity) and standardized abundance trends are the currency of:

- **Extinction-risk assessment** (IUCN Red List criteria tied to population reduction),
- **Data-limited fisheries methods** (life-history priors for $r$, $F_{MSY}$, generation time),
- **Macroecology** (allometric and life-history invariants across species).

Sharkipedia centralizes those measurements with traceable references. **sharkipediaR** lets you treat Sharkipedia like any other data API in R — filter, summarize, and plot with familiar **`%>%`** pipes and interactive **plotly** charts in this vignette.

Throughout this vignette we use **`example_carcharhinus()`** (blacknose shark, *Carcharhinus acronotus*), a data-rich coastal species with both traits and Northwest Atlantic trends. Live extraction is identical — swap in `sp_traits()` / `sp_trends()` when you are online.

---

## `sp_species()` — taxonomy for comparative frameworks

**File:** `R/sp_species.R`

`sp_species()` returns a **one-row tibble** with taxonomic ranks parsed from the species page header, plus provenance.

**Pipeline:** `resolve_species_url()` → `fetch_page()` → `parse_taxonomy()` → add `source_url` and `retrieved_at`.

```{r species-show}
meta
```

Use this to join trait or trend tables to order/family for comparative plots across many species (after a batch `sp_traits()` call).

**Live equivalent:**

```{r species-live, eval = FALSE}
sp_species("Carcharhinus acronotus")
```

---

## `sp_traits()` — life-history measurements in long format

**File:** `R/sp_traits.R`

`sp_traits()` is the workhorse for ecology. It returns one row per **measurement** with columns:

| Column | Meaning |
|--------|---------|
| `trait_group` | Sharkipedia grouping (Age, Length, Reproduction, …) |
| `trait_name` | Trait label (e.g. `Amat50`, `Linf`) |
| `value` | Reported value (character; coerce with `as.numeric()` when appropriate) |
| `standard` | Units or category (e.g. `Year`, `cm`) |
| `value_type` | Often `mean`, `median`, etc. |
| `sex` | `Male`, `Female`, `Pooled`, … |
| `location` | Geographic or study location |
| `reference` | Sharkipedia reference ID |
| `source_url`, `retrieved_at` | Reproducibility |

**Pipeline:** `fetch_page()` → `parse_traits_tables()` → `clean_traits()` → `validate_traits()`.

### How many traits, and from which groups?

```{r trait-inventory}
traits %>%
  count(trait_group, sort = TRUE)
```

### Plot 1 — Age at 50% maturity by sex

A classic input for generation length and demographic models: **age at 50% maturity (`Amat50`)** by sex.

```{r plot-amat50}
amat50 <- traits %>%
  filter(trait_name == "Amat50") %>%
  mutate(value_num = as.numeric(value))

p_amat50 <- ggplot(amat50, aes(
  x = sex, y = value_num, fill = sex,
  text = paste0(
    "Reference: ", reference,
    "<br>Location: ", location,
    "<br>Value: ", value_num, " years"
  )
)) +
  geom_col(width = 0.6, alpha = 0.85) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Age at 50% maturity — Carcharhinus acronotus",
    subtitle = "Hover for Sharkipedia reference and location",
    x = NULL,
    y = "Age (years)",
    fill = "Sex",
    caption = meta$source_url[[1]]
  ) +
  theme_shark() +
  theme(legend.position = "none")

as_interactive(p_amat50)
```

Each bar is traceable to a **`reference`** ID on Sharkipedia — exactly what you want in a methods paragraph.

### Plot 2 — Distribution of numeric traits across groups

```{r plot-trait-groups}
numeric_traits <- traits %>%
  mutate(value_num = suppressWarnings(as.numeric(value))) %>%
  filter(!is.na(value_num), trait_group %in% c("Age", "Length", "Reproduction"))

p_traits <- ggplot(numeric_traits, aes(
  x = trait_group, y = value_num, fill = trait_group,
  text = paste0(
    trait_name, " (", sex, ")",
    "<br>", value_num, " ", standard,
    "<br>", location
  )
)) +
  geom_boxplot(outlier.shape = 21, alpha = 0.7) +
  scale_y_continuous(trans = "log10", labels = scales::label_number()) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Numeric life-history measurements by trait group",
    subtitle = "Hover points for trait name, sex, and location",
    x = NULL,
    y = "Reported value (log scale)",
    fill = "Group",
    caption = "Example data: example_carcharhinus()"
  ) +
  theme_shark() +
  theme(legend.position = "none")

as_interactive(p_traits)
```

### dplyr workflow — matching the README blueprint

```{r dplyr-linf, eval = FALSE}
# With live data:
# traits_live <- sp_traits("Aetobatus narinari")

traits %>%
  filter(trait_name == "Linf") %>%
  summarise(
    n = n(),
    mean_linf = mean(as.numeric(value), na.rm = TRUE),
    .by = sex
  )
```

**Batch species:** pass a character vector; `sp_traits()` row-binds results and adds `species_input`.

```{r batch-traits, eval = FALSE}
sp_traits(c("Carcharhinus acronotus", "Alopias vulpinus"))
```

---

## `sp_trends()` — population trajectories for assessment & Red List thinking

**File:** `R/sp_trends.R`

Trend data are embedded in species pages as JSON inside `data-react-props` (server-rendered). `parse_trends_tables()` expands each trend to long format: **`year`**, **`value`**, **`location`**, **`unit`**, **`reference`**, **`trend_id`**.

**Pipeline:** `fetch_page()` → `parse_trends_tables()` → `clean_trends()` → `validate_trends()`.

```{r trends-preview}
trends %>%
  distinct(location, unit, reference, trend_id) %>%
  arrange(location)
```

### Plot 3 — Standardized index through time (single series)

Many Sharkipedia trends are **standardized indices** or relative abundances — ideal for showing directional change when absolute scale varies by survey.

```{r plot-one-trend}
one_trend <- trends %>%
  filter(trend_id == "3537")

p_trend <- ggplot(one_trend, aes(
  x = year, y = value,
  text = paste0(
    "Year: ", year,
    "<br>Index: ", round(value, 3),
    "<br>Reference: ", reference,
    "<br>", location
  )
)) +
  geom_line(linewidth = 0.9, colour = "#1f4e79") +
  geom_point(size = 1.8, colour = "#1f4e79") +
  labs(
    title = "Population trend — North Carolina to Brownsville (USA)",
    subtitle = "Hover points for year, index value, and reference",
    x = "Year",
    y = "Index value",
    caption = one_trend$trend_url[[1]]
  ) +
  theme_shark()

as_interactive(p_trend)
```

In a Red List or ERAs context, you would align such series to time windows required by criteria A2/A4, document the **`reference`**, and keep the **`retrieved_at`** stamp from `sp_trends()`.

### Plot 4 — Multiple locations, faceted

```{r plot-facet-trends}
trend_summary <- trends %>%
  group_by(location, unit, trend_id) %>%
  summarise(
    n_years = n(),
    year_min = min(year),
    year_max = max(year),
    .groups = "drop"
  ) %>%
  filter(n_years >= 10) %>%
  slice_head(n = 6)

plot_df <- trends %>%
  semi_join(trend_summary, by = "trend_id")

p_facet <- ggplot(plot_df, aes(
  x = year, y = value,
  colour = location,
  text = paste0(
    location,
    "<br>Year: ", year,
    "<br>Value: ", round(value, 3),
    "<br>Reference: ", reference,
    "<br>Unit: ", unit
  )
)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.2, alpha = 0.7) +
  labs(
    title = "Multiple abundance trends for one species",
    subtitle = "Interactive: zoom, pan, and hover for metadata",
    x = "Year",
    y = "Value (units vary by series)",
    colour = "Location",
    caption = meta$source_url[[1]]
  ) +
  theme_shark() +
  theme(legend.position = "bottom")

as_interactive(p_facet)
```

**Live extraction:**

```{r trends-live, eval = FALSE}
sp_trends("Carcharhinus acronotus")
```

---

## `sp_references()` — linking data to the literature graph

**File:** `R/sp_references.R`

Collects unique `/references/...` links from trait and trend tables on the species page.

```{r refs}
ex$references %>%
  slice_head(n = 8)
```

Join `traits$reference` to `references$reference_id` for bibliographic workflows (DOI lookup, `bibtex` keys).

---

## Discovery: `sp_species_urls()` and `sp_search()`

Before batch analysis, discover what Sharkipedia hosts:

```{r index-sample}
ex$species_index %>%
  slice_head(n = 6)
```

**`sp_species_urls()`** (`R/sp_species_urls.R`):

- Default: first index page only (polite).
- `all_pages = TRUE`: walks pagination (65+ pages) — use `max_pages` to cap load during development.

**`sp_search()`** (`R/sp_search.R`):

- Case-insensitive substring match on `species` or `slug`.
- Pass a pre-built `index` to avoid re-downloading the full catalogue.

```{r search-live, eval = FALSE}
idx <- sp_species_urls(all_pages = TRUE, max_pages = 5)
sp_search("Carcharhinus", index = idx)
```

---

## Reproducibility checklist (workshop-ready)

1. Record `packageVersion("sharkipediaR")` and `sessionInfo()`.
2. Store `source_url` and `retrieved_at` from every `sp_*()` call.
3. Archive raw outputs with `write_csv()` or `saveRDS()`.
4. Prefer building a species list once with `sp_species_urls()` rather than repeated index scrapes.

```{r reproducibility}
sessionInfo()
packageVersion("sharkipediaR")
```

---

## What is next in your analysis?

- Combine trait imputation with trends for data-poor relatives (Robin Hood / comparative approaches in the Dulvy tradition).
- Scale to many species with **conservative** rate limits (sequential `sp_traits()` loops; parallel only with care).
- Watch for Phase 4 functions (`sp_trait_catalogue()`, trait-centric queries) in future releases.

For implementation details of every internal function, see the **Architecture and function reference** vignette.
