---
title: "Ecological workflows"
author: "Pablo Fuenzalida"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ecological workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6,
  fig.align = "center"
)
```


## The story in one paragraph

[Sharkipedia](https://www.sharkipedia.org) links population time series to the papers they came from. **sharkipediaR** lets you pull those series into R and plot them like any other dataset — with provenance attached.

Here we focus on [Reid et al. (2011)](https://www.sharkipedia.org/references/reid2011) (*reid2011* on Sharkipedia): decadal trends from the New South Wales Shark Meshing Program ([DOI: 10.1071/MF10162](https://doi.org/10.1071/MF10162)).


## Load data

```{r load}
suppressPackageStartupMessages(library(tidyverse))
library(sharkipediaR)
library(ggplot2)
library(plotly)
library(viridis)
library(scico)

reid <- example_reid2011_trends()
white <- example_white_shark_trends()
```

Offline copies were built with `sp_trends()` on each species page. Online, the same code is:

```{r live-fetch, eval = FALSE}
species <- c("Carcharias taurus", "Carcharodon carcharias", "Galeocerdo cuvier",
               "Sphyrna zygaena", "Squatina australis")
reid_live <- map_dfr(species, sp_trends) %>% filter(reference == "reid2011")
```


## Prepare Reid (2011) NSW data

We express each series as **percent change from its first year** so different species can be compared on one axis.

```{r prep-reid}
species_labels <- c(
  "Carcharias taurus" = "Grey nurse",
  "Carcharodon carcharias" = "White shark",
  "Galeocerdo cuvier" = "Tiger shark",
  "Heterodontus portusjacksoni" = "Port Jackson shark",
  "Notorynchus cepedianus" = "Sevengill shark",
  "Sphyrna zygaena" = "Zebra shark",
  "Squatina australis" = "Angel shark"
)

reid_plot <- reid %>%
  group_by(species, trend_id) %>%
  arrange(year) %>%
  mutate(
    baseline = value[which.min(year)],
    pct_change = (value / baseline - 1) * 100
  ) %>%
  ungroup() %>%
  mutate(common_name = species_labels[species])
```


## Figure 1 — NSW meshing trends, all species

One panel per species (stacked vertically). Hover for year and percent change.

```{r fig-reid-facet}
p_reid <- ggplot(reid_plot, aes(x = year, y = pct_change, colour = common_name, group = common_name)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey55") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.2, alpha = 0.85) +
  facet_wrap(~ common_name, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
  scale_colour_viridis_d(option = "viridis", end = 0.9) +
  labs(
    title = "Relative change in meshing-program index — NSW",
    subtitle = "Reid et al. (2011); baseline = first year in each series",
    x = "Year",
    y = "Change from baseline (%)",
    caption = "Data: Sharkipedia reference reid2011 · sharkipediaR"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", hjust = 0)
  )

ggplotly(p_reid, tooltip = c("year", "pct_change", "common_name"))
```


## Figure 2 — White shark trends around the world

How does decline in NSW compare to other white-shark series on Sharkipedia? We pool every white-shark (`Carcharodon carcharias`) trend, again as **percent change from the first year** in each series.

```{r prep-white}
white_plot <- white %>%
  group_by(location, trend_id, reference) %>%
  arrange(year) %>%
  mutate(
    baseline = value[which.min(year)],
    pct_change = (value / baseline - 1) * 100
  ) %>%
  ungroup() %>%
  mutate(
    location_short = str_trunc(location, 40),
    is_nsw_reid = reference == "reid2011"
  )
```

```{r fig-white-facet}
p_white <- ggplot(
  white_plot,
  aes(x = year, y = pct_change, colour = is_nsw_reid, group = trend_id)
) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey55") +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  geom_point(size = 0.9, alpha = 0.75) +
  facet_wrap(~ location_short, ncol = 1, scales = "free_y") +
  scale_colour_manual(
    values = c("FALSE" = scico::scico(2, palette = "bilbao")[1],
               "TRUE" = scico::scico(2, palette = "bilbao")[2]),
    labels = c("Other sources", "Reid et al. (2011)"),
    name = NULL
  ) +
  labs(
    title = "White shark population trends — multiple populations",
    subtitle = "Highlighted series = Reid et al. (2011) NSW meshing program",
    x = "Year",
    y = "Change from baseline (%)",
    caption = "Data: Sharkipedia · sharkipediaR"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", hjust = 0),
    legend.position = "bottom"
  )

ggplotly(p_white, tooltip = c("year", "pct_change", "location", "reference"))
```


## Take-home

- **`sp_trends()`** turns Sharkipedia’s embedded charts into a long table (`year`, `value`, `location`, `reference`).
- Filtering by **`reference == "reid2011"`** recreates a published multi-species comparison in a few lines.
- **`ggplotly()`** adds exploration without extra data wrangling.

For function-by-function detail, see **Package architecture** under **Tutorials**.
