---
title: "Getting started with sharkipediaR"
author: "Pablo Fuenzalida"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with sharkipediaR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

## Why this package exists

[Sharkipedia](https://www.sharkipedia.org) is a curated, open database of shark and ray life-history traits and population abundance time series — the kind of comparative resource that underpins fisheries assessment, extinction-risk analysis, and macroecology ([Dulvy et al., 2022](https://www.nature.com/articles/s41597-022-01655-1)). **sharkipediaR** brings those data into R as tidy tibbles so your workflows look like any other modern ecological analysis: **`%>%`** pipes, explicit provenance, ggplot-ready tables, and interactive **plotly** charts in the ecological vignette.

The package is deliberately **not** a bulk mirror of the website. It is a polite, reproducible client: rate-limited requests, in-session caching, and HTML parsing separated from cleaning and validation.

## Installation

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ecologistpablo/SharkpediaR")
```

```{r load}
library(sharkipediaR)
library(dplyr)
```

## A thirty-second workflow

With an internet connection, you can pull data for a single species in three lines:

```{r live-demo, eval = FALSE}
meta   <- sp_species("Carcharhinus acronotus")
traits <- sp_traits("Carcharhinus acronotus")
trends <- sp_trends("Carcharhinus acronotus")
```

Every user-facing table includes **`source_url`** and **`retrieved_at`** so you can cite exactly what was downloaded and when — essential for Red List workshops, stock assessments, and supplementary materials.

## Offline example data

Vignettes and `R CMD check` use a frozen example dataset so documentation builds without hitting the live site:

```{r example}
ex <- example_carcharhinus()
names(ex)
```

| List element      | Description                                      |
|-------------------|--------------------------------------------------|
| `species_meta`    | Taxonomy + provenance (one row)                  |
| `traits`          | Long-format life-history trait measurements      |
| `trends`          | Long-format population trend observations        |
| `references`      | Reference IDs linked from the species page       |
| `species_index`   | Sample rows from the public species index        |

## Exported functions at a glance

| Function | Script | Purpose |
|----------|--------|---------|
| `sp_species_urls()` | `R/sp_species_urls.R` | Discover species names and URLs from the index |
| `sp_search()` | `R/sp_search.R` | Filter the index by partial scientific name |
| `sp_species()` | `R/sp_species.R` | Taxonomy metadata for one species |
| `sp_traits()` | `R/sp_traits.R` | Life-history traits (long format) |
| `sp_trends()` | `R/sp_trends.R` | Population trends (long format) |
| `sp_references()` | `R/sp_references.R` | Bibliographic reference links |
| `fetch_page()` | `R/fetch.R` | Low-level HTML retrieval (advanced) |
| `example_carcharhinus()` | `R/example_data.R` | Offline example data |

Internal parsing and cleaning live in `R/parse.R`, `R/clean.R`, and `R/validate.R` — see the **Architecture and function reference** vignette.

## Where to go next

- **Ecological workflows** — trait summaries, interactive **plotly** trend plots, `%>%` workflows.
- **Architecture and function reference** — how every function in each script works, with pipeline examples.

## Building the pkgdown site

```{r pkgdown, eval = FALSE}
install.packages(c("pkgdown", "ggplot2", "plotly", "htmltools"))
pkgdown::build_site()
```

The site will include these vignettes under **Articles** automatically.
