---
title: "Getting started"
author: "Pablo Fuenzalida"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```


## What this tutorial is for

This page is a **quick tour** of **sharkipediaR** — not the full package manual (that lives on the home page and in **Reference**).

[Sharkipedia](https://www.sharkipedia.org) holds published shark and ray **traits** and **population trends**. This package pulls that into R as tidy tables you can filter and plot.


## Load the package

```{r load}
suppressPackageStartupMessages(library(tidyverse))
library(sharkipediaR)
```


## Three commands worth knowing

When you are online, almost everything starts here:

```{r live, eval = FALSE}
sp_species("Carcharodon carcharias")   # taxonomy
sp_traits("Carcharodon carcharias")    # life-history table
sp_trends("Carcharodon carcharias")      # abundance through time
```

Each result includes **`source_url`** and **`retrieved_at`** — keep them for methods sections and supplementary files.


## Try it without the internet

Vignettes use frozen data so GitHub Actions can build the site:

```{r offline}
ex <- example_carcharhinus()
glimpse(ex$traits)
```

For trend examples in the next tutorial, see `example_reid2011_trends()` and `example_white_shark_trends()`.


## Where to go next

**Ecological workflows** — ggplot + plotly, including [Reid et al. (2011)](https://www.sharkipedia.org/references/reid2011) NSW meshing trends.


**Package architecture** — plain-language guide to every function.


## Install (if you have not yet)

```{r install, eval = FALSE}
devtools::install_github("ecologistpablo/SharkipediaR")
```
